Learning with misspecified models: the case of overestimation

### 2023-10-14 Paper/draft.tex
recompile draft

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Learning with misspecified models: the case of overestimation},
  pdfauthor={Jimena Galindo},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=2.5cm]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

@@

\newcommand{\EE}[1]{\mathbb{E}\left[#1\right]}
\newcommand{\PP}[1]{\mathbb{P}\left(#1\right)}
\newcommand{\CP}[2]{\mathbb{P}\left(#1 \,| \, #2 \right)}
\newcommand{\CE}[2]{\mathbb{E}\left[#1\,|\,#2 \right]}
\usepackage[bottom]{footmisc}
\usepackage[doublespacing]{setspace}
\usepackage[normal]{caption}
\usepackage{dsfont}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{hyperref}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}

\title{Learning with misspecified models: the case of overestimation}
\author{Jimena Galindo}
\date{October 13, 2023}

\begin{document}
\maketitle
\begin{abstract}
I design a framework and laboratory experiment that allows for the
comparison of multiple theories of misspecified learning. I focus on a
framework with endogenous information and two complementary
fundamentals: an ego-relevant parameter and a state. Within this
framework I study three forces that can lead to misspecified beliefs:
dogmatism, learning traps and biased updating. I find that biased
updating is the main driver of misspecifications. In addition, I vary
the degree of ego-relevance of the parameter by introducing a stereotype
treatment. The data is consistent with biased updating in both cases but
for potentially different reasons: subjects attribute successes to their
own ability and failures to luck. In the stereotype treatment they
over-compensate for initial negative biases by over-attributing positive
signals to the ability of others. This translates into similar observed
choices but different dynamisc in beliefs.
\end{abstract}

@@

A growing body of literature in economics explores how people develop
incorrect beliefs about fundamentals. Most of this research centers on
scenarios where agents passively observe the world, and incorrectly
integrate the received information into their
beliefs\footnote{See \citet{benjamin2019} for a review of the literature on errors of 
probabilistic thinking}. However, many real-world situations cast agents
as active participants in the generation of information. In these cases,
the information they acquire is influenced by their actions and their
subsequent behavior is in turn determined by how they incorporate the
information into their beliefs. This further affects the information
generating process.

Consider a student who needs to decide how much effort to put into
studying for an exam. Their decision will depend on two different
factors: their belief about their intrinsic ability, and their belief
about how difficult the exam will be. The outcome that they observe will
be affected by how much they decide to study. Imagine that the student
puts in a moderate amout of effort and gets a surprisingly good grade.
Did they get a good grade because they are smarter than they thought? Or
because the exam was easier than they had anticipated? Their future exam
preparation strategy will depend fundamentally on which line of
reasoning they take; in turn, their performance in the next exam will
further affect the outcome they observe. This type of feedback loop is
what I call and \emph{endogenous information 
setting} and is at the center of the forces that I study.

I compare a set of theories that model learning in an endogenous
information setting and which can rationalize the persistence of
misspecified beliefs. To do so, I first come up with a unifying
framework in which each of the theories makes different predictions. I
then implement this framework in a laboratory experiment. The framework
that I use features an agent who, in order to make optimal decisions,
needs to learn a parameter that regards their own characteristics (an
\emph{ego-relevant} parameter), as well as an exogenously determined
state---in the case of the student it would be their intrinsic ability
and the difficulty of the exam.

Misspecified beliefs about an ego-relevant parameter are often referred
to as overconfidence and have been documented by behavioral
science\footnote{\citet{kelley1980} 
provides a review of the psychology literature.} and
economics\footnote{See \citet{benjamin2019} 
for a review of the literature in economics.} in a variety of settings.
\citet{Oster2013} find that subjects who are at high risk of having
Huntington's disease overestimate their probability of being healthy and
make retirement decisions as if they were. \citet{Hoffman2020} show that
workers overestimate the quality of their match to their current
employment and are unlikely to look for other opportunities.
\citet{Camerer1999} find that entrepreneurs are overconfident about the
quality of their enterprise, which leads to excessive entry and early
exit from markets.

@@

These theories provide the main building blocks for a simplified
framework that can be directly implemented in a laboratory experiment.
The goal is primarily to identify the which of the 3 forces---dogmatism,
the presence of learning traps or missatribution bias---better explains
the observed behavior. To do so, I compare the behavior predicted by
each theory to the benchmark given by the fully Bayesian updating
procedure.

On a secondary level, I also aim to understand if the misspecification
is inherently linked to the ego-relevance of one of the parameters or if
it is a more general phenomenon. To do so I vary the degree of
ego-relevance by introducing a treatment in which subjects learn about
the ability of another participant. In this treatment, the participants
know only the gender and nationality of the other, which can induce
stereotypes---a different type of misspecification.\\
If correct learning about the parameters happens at higher rates in the
stereotype treatment, it would suggest that the bias is intrinsically
linked to the ego-relevance of the parameter. In contrast, if similar
biased behavior arises in both treatments, it is more likely that the
main driver of these types of misspecified beliefs is the updating
procedure itself or the endogenous information process.

I find that, although some agents do fall into learning traps, the
behavior of most subjects is better explained by misattribution bias. I
also find that misattribution is no more prevalent in the ego-relevant
than in the stereotype condition. This suggests that the main driver of
the misspecification is the updating proceduce. However, the underlying
mechanism by which the bias is generated may be different in both
treatments: While in the ego-relevant condition, the intuitive
explanation behind the bias is that the agent prefers to hold themselves
in high esteem, in the stereotype condition it seems to be driven by
some sort of self-censoring---when subjects realize that they
underestimated the ability of another participant based on their gender
and nationality, they over-compensate by overestimating their ability.

Finally, I structurally estimate the parameters of the models to study
heterogeneity in the sample. I find that even at an individual level the
behavior is better explained by misattribution bias for most subjects.
There is a smaller group of subjects that can be better explained
through dogmatic beliefs and hypothesis testing and none of them behave
in line with the fully Bayesian benchmark.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=2.5cm]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

@@

\newcommand{\EE}[1]{\mathbb{E}\left[#1\right]}
\newcommand{\PP}[1]{\mathbb{P}\left(#1\right)}
\newcommand{\CP}[2]{\mathbb{P}\left(#1 \,| \, #2 \right)}
\newcommand{\CE}[2]{\mathbb{E}\left[#1\,|\,#2 \right]}
\usepackage[bottom]{footmisc}
\usepackage[doublespacing]{setspace}
\usepackage[normal]{caption}
\usepackage{dsfont}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{hyperref}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Learning with misspecified models: the case of overestimation},
  pdfauthor={Jimena Galindo},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Learning with misspecified models: the case of overestimation}
\author{Jimena Galindo}
\date{October 14, 2023}

\begin{document}
\maketitle
\begin{abstract}
I design a framework and laboratory experiment that allows for the
comparison of multiple theories of misspecified learning. I focus on a
framework with endogenous information and two a data generating process
ruled by two fundamentals: an ego-relevant parameter and a state. Within
this framework I study three forces that can lead to misspecified
beliefs: initial misspecifications, learning traps and biased updating.
I find that biased updating is the main driver of misspecifications. In
addition, I vary the degree of ego-relevance of the parameter by
introducing a stereotype treatment. The data is consistent with biased
updating in both cases but for potentially different reasons: In the
ego-relevant treatment subjects attribute successes to their own ability
and failures to luck; in the stereotype treatment they over-compensate
for initial negative biases by over-attributing positive signals to the
ability of others. This translates into similar observed choices but
different dynamic in beliefs.
\end{abstract}

@@

A growing body of literature in economics explores how people develop
incorrect beliefs about fundamentals. Most of this research centers on
scenarios where agents passively observe the world, and incorrectly
integrate the received information into their
beliefs.\footnote{See \citet{benjamin2019} for a review of the literature on errors of 
probabilistic thinking} However, many real-world situations cast agents
as active participants in the generation of information. In these cases,
the information they acquire is influenced by their actions and
subsequent behavior is in turn determined by how they incorporate the
information into their beliefs.

Consider a student who needs to decide how much effort to put into
studying for an exam. Their decision will depend on two factors: their
belief about their intrinsic ability, and their belief about how
difficult the exam will be. The outcome they observe will be affected by
how much they decide to study. Imagine that the student puts in a
moderate amout of effort and gets a surprisingly good grade. Did they
get a good grade because they are smarter than they thought? Or because
the exam was easier than they had anticipated? Their future exam
preparation strategy will depend fundamentally on which line of
reasoning they take. This feedback loop is refered to as an
\emph{endogenous information 
process}.

I compare a set of theories that model learning in an endogenous
information setting and which can rationalize the persistence of
misspecified beliefs. I describe a unifying framework that nests
multiple theories of model learning and generates testable predictions
for each of them. I cast this framework in a laboratory experiment
design and test these predictions to identify which of these theories
are consistent with the behavior I observe in the lab. The framework
features an agent who needs to learn two parameters: one that regards
their own characteristics (an \emph{ego-relevant} parameter), and an
exogenously determined state. In the context of the student preparing
for exams, the two parameters would be their intrinsic ability and the
difficulty of the exam.

Misspecified beliefs about an ego-relevant parameter are often referred
to as overconfidence (or underconfidence) and have been documented by
behavioral scientists\footnote{\citet{kelley1980} 
provides a review of the psychology literature.} and
economists\footnote{See \citet{benjamin2019} 
for a review of the literature in economics.} in a variety of settings.
\citet{Oster2013} find that subjects who are at high risk of having
Huntington's disease overestimate their probability of being healthy and
make retirement decisions as if they were healthy. \citet{Hoffman2020}
show that workers overestimate the quality of their match to their
current employment and are unlikely to look for other opportunities.
\citet{Camerer1999} find that entrepreneurs are overconfident about the
quality of their enterprise, which leads to excessive entry and early
exit from markets.

@@

These theories provide the main building blocks for a simplified
framework that can be directly implemented in a laboratory experiment.
The goal is to identify the which of the 3 forces---the presence of
learning traps, mispecified initial beliefs, or missatribution
bias---better explains the observed behavior. To determine the fit of
the models, I compare the behavior predicted by each theory to the
benchmark given by the fully Bayesian updating procedure.

I also study whether the misspecification is inherently linked to the
ego-relevance the parameters or a more general phenomenon. I vary the
degree of ego-relevance by introducing a treatment in which subjects
learn about the ability of another participant. In this treatment, the
participants know only the gender and nationality of the other, which
can induce stereotypes---a different type of misspecification.\\
If correct learning about the parameters happens at higher rates in the
stereotype treatment, it would suggest that the bias is intrinsically
linked to the ego-relevance of the parameter. In contrast, if similar
biased behavior arises in both treatments, it is more likely that the
main driver of these types of misspecified beliefs is the updating
procedure itself or the endogenous information process.

Although some agents do fall into learning traps, I find that the
behavior of most subjects is better explained by misattribution bias. I
also find that misattribution is no more prevalent in the ego-relevant
than in the stereotype condition. This suggests that the main driver of
the misspecification is the updating proceduce. However, the underlying
mechanism by which the bias is generated may be different in both
treatments: While in the ego-relevant condition, the intuitive
explanation behind the bias is that the agent prefers to hold themselves
in high esteem, in the stereotype condition it seems to be driven by
some sort of self-censoring---when subjects realize that they
underestimated the ability of another participant based on their gender
and nationality, they over-compensate by overestimating their ability.

Finally, I estimate the structural parameters of the models to study
model-heterogeneity in the sample. I find that even at an individual
level the behavior is better explained by misattribution bias for most
subjects. There is a smaller group of subjects that can be better
explained through dogmatic beliefs and hypothesis testing and none of
them behave in line with the fully Bayesian benchmark.
